A mixture of hidden Markov models to predict the lymphatic spread in head and neck cancer depending on primary tumor location

### 2024-10-03 manuscript.qmd
Updates from Overleaf
reworked the math again to only us k

$$
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \left[P \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}, T_j \right) \right]^{N_{ij}}
$$ {#eq-likelihood-HMM}

In @eq-likelihood-HMM we take the likelihood of observing a patient in t-stage $T_j$ and with involvement $\xi_i$ to the power of $N_{ij}$, corresponding to the number of observations of the patients in state $i$ and t-stage $T_j$ (either early or late) in the dataset $\mathbf{D}$.

@@

$$
\sum_{m=1}^M \pi_m^s = 1, \quad \forall s
$$

@@

The latent variables $\boldsymbol{\epsilon}_{k}$, on the other hand, are patient-specific and represent the assignment of each patient to a particular model. We will later generalize patients with the same diagnosis to the latent variable $\boldsymbol{\epsilon}_{i,j}$. The subscripts $i$ and $j$ represent that the latent variables are only dependent on the diagnosis of the patient, which consists of lymphatic involvement $\boldsymbol{\xi}_i$ and t-stage $T_j$. The latent variables are central to the model, as they are hidden (unobserved) and must be inferred.

@@

$$
P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k=1}^{K} \prod_{m=1}^{M} \left[ \pi_m^s P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
$${#eq-complete-likelihood}

@@

* $\pi_m^s$ is the mixing parameter for subsite $s$ and model $m$,
* $P(\mathbf{D}_k | \boldsymbol{\theta}_m)$ is the likelihood of patient $k$'s data given that it was generated by model $m$,
* $\boldsymbol{\theta}_m$ represents the parameters of model $m$,
* $\epsilon_{k}^m$ is the latent variable that indicates whether patient $k$ was generated by model $m$.

@@

We can now apply the mixture model to our dataset $\mathbf{D}$ comprising the number of patients $N_{ijs}$ diagnosed with LNL involvement state $\mathbf{X} = \boldsymbol{\xi}_i$, t-stage $T_j$, with a primary tumor in subsite $s$.

@@

$$
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_i \prod_j \prod_s \left[ \prod_{m=1}^M \left[ \pi_m^s P_m \left( \mathbf{X}=\boldsymbol{\xi}_i \mid \boldsymbol{\theta}_m, T_j \right) \right]^{\gamma(\epsilon_{i,j}^m)} \right]^{N_{ijs}}
$$ {#eq-complete-likelihood-explicite}

In @eq-complete-likelihood we compute the likelihood of observing state $\boldsymbol{\xi}_i$ with t-stage $T_j$ under model $m$ similarly to @eq-likelihood-HMM. However, we weight each model with the mixing parameter $\pi_m^s$. Since the mixing parameters are subsite dependent, we differ between each subsite.

@@

In the E-step, we compute the expected value of the latent variables $\boldsymbol{\epsilon}$, which represent the probabilities of a patient $k$ originating from one of the models $m$. Given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, the expectation value $\gamma(\epsilon_m^k)$ is calculated as:

$$
\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_{i,j}) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j)}{\sum_{n=1}^M \pi_n^s P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_n}, T_j)}
$${#eq-expectation-step}

THE OTHER OPTION

$$
\mathbb{E}[\epsilon^m] = \gamma(\epsilon^m_k) = \frac{\pi_m^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{j=1}^M \pi_j^s P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_j}, T_k)}
$${#eq-expectation-step2}

Here, the mixing parameter $\pi_m^s$ depends on the subsite location of the primary tumor of patient $k$.

@@

$$
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_{i} \sum_j \sum_{s} N_{i,j,s} \sum_{m=1}^M \gamma(\epsilon^m_{i,j}) \left( \ln \pi_m^s + \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \boldsymbol{\theta_m}, T_j) \right)
$${#eq-maximization-step}

@@

$$
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{i} \sum_j \gamma(\epsilon^m_{i,j}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_j)
$$

>>>>

$$
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[P \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}, T_k \right) \right]
$$ {#eq-likelihood-HMM}

In @eq-likelihood-HMM we compute the likelihood of observing patient the diagnosis of each patient $k$, i.e. t-stage $T_k$ and involvement $\xi_k$.

@@

$$
\sum_{m}^M \pi_m^s = 1, \quad \forall s
$$

@@

The latent variables $\boldsymbol{\epsilon}_{k}$, on the other hand, are patient-specific and represent the assignment of each patient to a particular model. The latent variables are central to the model, as they are hidden (unobserved) and must be inferred.

@@

$$
P(\mathbf{D}, \boldsymbol{\epsilon} | \boldsymbol{\theta}, \boldsymbol{\pi}) = \prod_{k}^{K} \prod_{m}^{M} \left[ \pi_m^{s_k} P(\mathbf{D}_k | \boldsymbol{\theta}_m, \boldsymbol{\pi}) \right]^{\epsilon_{k}^m}
$${#eq-complete-likelihood}

@@

* $\pi_m^{s_k}$ is the mixing parameter for patient $k$ with a primary tumor in subsite $s_k$ and model $m$,
* $P(\mathbf{D}_k | \boldsymbol{\theta}_m)$ is the likelihood of patient $k$'s data given that it was generated by model $m$,
* $\boldsymbol{\theta}_m$ represents the parameters of model $m$,
* $\epsilon_{k}^m$ is the latent variable that indicates whether patient $k$ was generated by model $m$.

@@

@@

$$
P \left( \mathbf{D} \mid \boldsymbol{\theta}, \boldsymbol{\pi}\right) = \prod_k^K \left[ \prod_{m}^M \pi_m^{s_k} P_m \left( \mathbf{X}=\boldsymbol{\xi}_k \mid \boldsymbol{\theta}_m, T_k \right) \right]^{\gamma(\epsilon_k^m)} 
$$ {#eq-complete-likelihood-explicite}

In @eq-complete-likelihood we compute the likelihood of observing the patient's state $\boldsymbol{\xi}_k$ with t-stage $T_k$ under model $m$ similarly to @eq-likelihood-HMM. However, we weight each model with the mixing parameter $\pi_m^{s_k}$. Since the mixing parameters are subsite dependent, we differ between each subsite.

@@

In the E-step, we compute the expected value of the latent variables $\boldsymbol{\epsilon}$, which represent the probabilities of a patient $k$ originating from one of the models $m$. Given the current estimates of $\boldsymbol{\theta}$ and $\boldsymbol{\pi}$, the expectation value $\gamma(\epsilon_k^m)$ is calculated as:

$$
\mathbb{E}[\epsilon^m_k] = \gamma(\epsilon^m_{k}) = \frac{\pi_m^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k)}{\sum_{n}^M \pi_n^{s_k} P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_n}, T_k)}
$${#eq-expectation-step}

@@

$$
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}, \boldsymbol{\pi}) = \sum_k^K \sum_{m}^M \gamma(\epsilon^m_{k}) \left( \ln \pi_m^{s_k} + \ln P(\mathbf{X} = \boldsymbol{\xi}_k \mid \boldsymbol{\theta_m}, T_k) \right)
$${#eq-maximization-step}

@@

$$
\ln P(\mathbf{D}, \boldsymbol{\epsilon} \mid \boldsymbol{\theta}_m) = \sum_{k}^K \gamma(\epsilon^m_{k}) \ln P(\mathbf{X} = \boldsymbol{\xi}_i \mid \theta_m, T_k)
$$
